Demographic inference using the SFS with \moments and \demes

### 2025-08-25 chapter/paper.tex
Added a null model, better SFS parsing scripts and edited manuscript

\usepackage{authblk}
\title{Demographic inference using the SFS with \moments and \demes}
\author[1]{Nick Collier}
\author[1,*]{Aaron P. Ragsdale}
\affil[1]{Department of Integrative Biology, University of Wisconsin--Madison}
\affil[*]{apragsdale@wisc.edu}
\affil[ ]{ORCID IDs: 0009-0005-0385-9798 (NC), 0000-0003-0715-3432 (APR)}

\usepackage[margin=1in, headheight=15pt, headsep=10pt]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{N Collier and AP Ragsdale}
\rhead{Demographic inference with \moments and \demes}

@@

Below, we show two examples of multi-population demographic inference using the
joint SFS. The first example simulates data with known demographic parameters,
which we try to reinfer using both the original model and simpler
misspecified models. In the second example, we infer a relatively simple
human-Neanderthal history using data from two human populations and a
high-coverage ancient sample from the Neanderthal lineage. For both of these
examples, we briefly describe data processing and inference setup, emphasizing
the high-level components of the analyses. For detailed information about each
example, including managing data, specifying models, performing inference,
computing confidence intervals and visualizing fits, we refer readers to the
accompanying GitHub repository
(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
is maintained with up-to-date versioning.

@@

\subsection*{Example 2: Inferring human-Neanderthal demographic parameters}

This example uses publicly available data to infer a demographic model for two
human groups (a northern European and western African population) and a
Neanderthal individual. As shown below, even with just a single sample from the
Neanderthal lineage, we are able to recover key parameters such as the
human-Neanderthal divergence and introgression proportion.

We obtained modern human genome sequences from a recent resequencing of the 
1000 Genomes Project cohort \cite{byrska2022high}, and used the high-coverage 
sequence of the Vindija Neandertal from \cite{prufer2017high}. 
From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
GBR (British from England and Scotland) sequences as our modern human samples.  
We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
were not genotyped in the high-coverage archaic genome sequences, 
that lacked high-confidence ancestral state assignments, or fell within 10 kb of 
exonic or promoter regions. This left us with approximately 960 Mb of 
well-characterized and putatively neutrally-evolving sequence.
We estimated the genome-wide SFS using built-in \moments parsing functions.
Details concerning data processing (including our liftover of the Vindija 
sequence to a more recent genome build) can be found in the linked GitHub 
repository.

The model that we wished to fit is relatively simple but still has many
parameters ($11$ in the example we present), so we constructed it in stages by
first fitting simpler one- and two-population models.  Throughout, we used
$u=1.5\times 10^{-8}$ as an estimate of the mutation rate per generation and 29
years as the average generation time. When the goal is to infer a complex
demographic history involving multiple populations, we typically start simple,
with subsets of the full data, and sequentially add complexity.  Here, we began
by fitting one-population models with ancestral and recent size changes for MSL
and GBR.  For the MSL model, we allowed three epochs with constant effective
size in each epoch. We attributed to GBR a sharp contraction $\sim$60 kya
followed by exponential growth to represent the out-of-Africa bottleneck and
subsequent rapid expansion. YAML files representing the models and parameters
in this section can be found in the linked GitHub repository.

\reviewercomment{maybe it would be helpful to also include the likelihood (or AIC?) of each step to show how it changes with the addition of further parameters, and how it can be used to inform the models design}
\aprcomment{this suggestion would only make sense for models fit to the same data SFS, but it could be done to compare the fit without vs. with the Neanderthal pulse. Maybe we could demonstrate the likelihood ratio test, if it works? what do you think, Nick?}

We next stitched these one-population models together and refitted the 
resulting demographic model to get a two-population model 
that included continuous symmetric 
migration between MSL and GBR and an ancestral size expansion. Following this,
we added a Neandertal branch and fitted the 
Neandertal-modern human divergence time, the Neandertal effective size, and the 
Neandertal-to-GBR admixture proportion. We also fitted the divergence time 
between the sampled Vindija Neandertal lineage and the Neandertal lineage which 
admixed with modern non-African humans. Finding that this parameter tended to 
converge to its lower bound (imposed by the estimated time of death of the 
Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature 
(90 kya, \cite{prufer2017high}) in a second round of optimization.
Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
Fixing parameters for which good estimates are available in the literature 
(see, e.g., \cite{sumer2025earliest}) is often reasonable in the first stages of 
model optimization, and we expected this value to be especially difficult to fit 
due to the small pulse proportion and Neandertal sample size. In the second 
round of optimization on the three-population model, we refit all parameters 
aside from the two event times mentioned above to refine the estimates.

After this refitting step, we were left with parameters in good agreement with 
prior authors, with a Neandertal-modern human divergence time of 547 kya, 
MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 
percent. We observe some large, systematic residuals in our best-fit model 
(Figure~\ref{fig:humans}G-I), suggesting model misspecification that could be
improved through further exploration of model space.

@@

\begin{figure}
    %\includegraphics{../example2/fig2.pdf}
    \caption{
        Inference using human and Neanderthal data.
        (A) The best-fit model including the Vindija Neanderthal and two human
        populations.
        (B, C) Marginal SFS fits between the model and empirical data are shown
        for MSL and GBR.
        (D--I) Marginal two-way spectra predicted by the model and their residuals 
        against the data are plotted for the three pairs of populations MSL, GBR 
        (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
    }
    \label{fig:humans}
\end{figure}

>>>>

\usepackage{authblk}
\title{Demographic inference using the SFS with \moments and \demes}
\author[1]{Nicholas W. Collier}
\author[1,*]{Aaron P. Ragsdale}
\affil[1]{Department of Integrative Biology, University of Wisconsin--Madison}
\affil[*]{apragsdale@wisc.edu}
\affil[ ]{ORCID IDs: 0009-0005-0385-9798 (NWC), 0000-0003-0715-3432 (APR)}

\usepackage[margin=1in, headheight=15pt, headsep=10pt]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{NW Collier and AP Ragsdale}
\rhead{Demographic inference with \moments and \demes}

@@

Below, we show two examples of multi-population demographic inference using the
joint SFS. The first example simulates data with known demographic parameters,
which we try to reinfer using both the original model and simpler
misspecified models. In the second example, we infer a relatively simple
human-Neandertal history using data from two human populations and a
high-coverage ancient sample from the Neandertal lineage. For both of these
examples, we briefly describe data processing and inference setup, emphasizing
the high-level components of the analyses. For detailed information about each
example, including managing data, specifying models, performing inference,
computing confidence intervals and visualizing fits, we refer readers to the
accompanying GitHub repository
(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
is maintained with up-to-date versioning.

@@

\subsection*{Example 2: Inferring human-Neandertal demographic parameters}

This example uses publicly available data to infer a demographic model for two
modern human populations (a northern European and western African population) 
and a Neandertal individual. As shown below, even with just a single sample 
from the Neandertal lineage, we are able to recover key parameters such as the
human-Neandertal divergence time and fraction of ancestry contributed to modern 
Europeans by Neandertals.

We obtained modern human genome sequences from a recent resequencing of the 
1000 Genomes Project cohort \cite{byrska2022high}, and used the high-coverage 
sequence of the Vindija Neandertal genome published by \cite{prufer2017high}. 
From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
GBR (British from England and Scotland) sequences as our modern human samples.  
We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
were not genotyped in the high-coverage archaic genome sequences, 
that lacked high-confidence ancestral state assignments, or fell within 10 kb of 
exonic or promoter regions. This left us with approximately 960 Mb of 
well-characterized and putatively neutrally-evolving sequence.
We estimated the genome-wide SFS using built-in \moments parsing functions.
Details concerning data processing (including our liftover of the Vindija 
sequence to a more recent genome build) can be found in the linked GitHub 
repository.

The model that we wished to fit is relatively simple but still has many
parameters ($11$ in the example we present), so we constructed it in stages by
first fitting marginal one- and two-population models. When our goal is to infer 
a complex demographic history involving multiple populations, it is helpful to 
start simple, with subsets of the full data, and sequentially add complexity.  
Here, we began by fitting one-population models with ancestral and recent size 
changes for the MSL and GBR populations.  For the MSL model, we allowed three 
epochs with constant effective size in each epoch. We attributed to GBR a sharp 
contraction $\sim$60 kya followed by exponential growth to represent the 
out-of-Africa bottleneck and subsequent rapid population expansion.  Throughout, 
we used $u=1.5\times 10^{-8}$ as an estimate of the mutation rate per generation 
and 29 years as the average generation time. YAML files representing the models 
and parameters in this section can be found in the linked GitHub repository.

We next stitched these one-population models together and refitted the resulting 
demographic model to get a two-population model that included continuous 
symmetric migration between MSL and GBR and an ancestral size expansion. 
Following this, we added a Neandertal branch and fitted the 
Neandertal-modern human divergence time, the Neandertal effective size, and the 
Neandertal-GBR pulse proportion, leaving other parameters fixed. We also 
fitted the divergence time between the sampled Vindija Neandertal lineage and 
the Neandertal lineage which admixed with modern non-African humans. Finding 
that this parameter tended to converge to its lower bound (imposed by the 
estimated time of death of the Vindija Neandertal, ~55 kya), we fixed it to a 
value supported by literature (90 kya, \cite{prufer2017high}) in a second round 
of optimization. Rather than fitting it, we fixed the time of the Neandertal-GBR 
pulse to 50 kya. Fixing parameters for which good estimates are available in the 
literature  (see, e.g., \cite{sumer2025earliest}) is often reasonable in the 
first stages of model optimization, and we expected this value to be especially 
poorly constrained due to the small pulse proportion and Neandertal sample size. 
In the second round of optimization on the three-population model, we refit all 
parameters aside from the two event times mentioned above to refine the 
estimates.

After this refitting step, we were left with parameters in good agreement with 
those inferred by prior authors, with a Neandertal-modern human divergence time 
of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion 
of 2.7 percent. We observe some large, systematic residuals in our best-fit model 
(Figure~\ref{fig:humans}G-I), suggesting model misspecification that could be
improved through further exploration of model space.

We also fitted a nested model without the Neanderthal-GBR pulse, which had a 
substantially lower likelihood than the pulse model (log-likelihood [LL] 
$\approx -127,552$, against LL $\approx -25,103$ for the pulse model). We did 
not perform a formal hypothesis test; the null distribution of the 
likelihood-ratio test (LRT) statistic is overdispersed under composite 
likelihood, inflating the rate of false positives and rendering its use 
unjustifiable in this setting. \moments does incorporate a function 
(\texttt{moments.Godambe.LRT\_adjust}) which computes a multiplicative
adjustment, intended to restore the expected null distribution, using 
first-order moment matching \cite{coffman2016computationally}. This method 
requires bootstrapping over the data. Otherwise, we could conduct a hypothesis 
test by fitting the nested and pulse models to a large number of bootstrap 
samples and calculating the rate of rejection across replicates-- which has the 
benefit of producing a distribution of LRT statistics, but is time-consuming
and computationally expensive.

@@

\begin{figure}
    %\includegraphics{../example2/fig2.pdf}
    \caption{
        Inference using human and Neandertal data.
        (A) The best-fit model including the Vindija Neandertal and two human
        populations.
        (B, C) Marginal SFS fits between the model and empirical data are shown
        for MSL and GBR.
        (D--I) Marginal two-way spectra predicted by the model and their residuals 
        against the data are plotted for the three pairs of populations MSL, GBR 
        (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
    }
    \label{fig:humans}
\end{figure}
